title: "Understanding Semivariograms"
author: "Your Name"
date: "`r Sys.Date()`"
output: github_document
---

# Introduction to Semivariograms
A semivariogram describes how data values change with increasing distance. It helps quantify **spatial dependence**.

## Spatial Dependence and Tobler's Law
Tobler’s First Law of Geography states:
> "Everything is related to everything else, but near things are more related than distant things."

In semivariograms:
- **Low distances** → Similar values (low variance).
- **Large distances** → Increasing dissimilarity.

---

# **Semivariogram Components**
The key parameters of a semivariogram are:

- **Nugget (`C₀`)**: Variance at distance = 0 (measurement error or microscale variation).
- **Sill (`C`)**: Total variance when the semivariogram levels off.
- **Range (`A`)**: Distance at which spatial autocorrelation disappears.

### **Plotting a Sample Semivariogram in R**
```{r semivariogram-plot, echo=FALSE}
library(ggplot2)

# Define parameters
distances <- seq(0, 100, by=5)  # Distance values
nugget <- 15  # Nugget above zero
sill <- 160  # Sill where semivariance levels off
range_value <- 30  # Range where spatial dependence disappears

# Spherical model approximation
spherical_variogram <- function(h, nugget, sill, range_value) {
  ifelse(h < range_value, 
         nugget + (sill - nugget) * (1.5 * (h / range_value) - 0.5 * (h / range_value)^3), 
         sill)
}

# Compute semivariance values
semivariance <- sapply(distances, spherical_variogram, nugget = nugget, sill = sill, range_value = range_value)

# Create a data frame
data <- data.frame(Distance = distances, Semivariance = semivariance)

# Plot using ggplot2
ggplot(data, aes(x = Distance, y = Semivariance)) +
  geom_line(color = "blue", linewidth = 1.2) +  # Semivariogram line
  geom_hline(yintercept = sill, linetype = "dashed", color = "red") +  # Sill
  geom_hline(yintercept = nugget, linetype = "dashed", color = "green") +  # Nugget
  geom_vline(xintercept = range_value, linetype = "dashed", color = "purple") +  # Range
  annotate("text", x = 5, y = nugget + 5, label = "Nugget", color = "green", fontface = "bold") +
  annotate("text", x = 80, y = sill - 5, label = "Sill", color = "red", fontface = "bold") +
  annotate("text", x = range_value + 2, y = sill / 2, label = "Range", color = "purple", fontface = "bold", angle = 90) +
  labs(title = "Improved Semivariogram Diagram", x = "Distance", y = "Semivariance") +
  ylim(0, 170) +  # Set y-axis range from 0 to 170
  theme_minimal()
```

---

# **Working Through a Real Example: The Meuse Dataset**
```{r empirical-variogram, message=FALSE}
library(sp)
library(gstat)
library(gridExtra)

data(meuse, package = "sp")
coordinates(meuse) <- ~x+y  # Define spatial coordinates

# Compute the Empirical Variogram
variogram_model <- variogram(zinc ~ 1, meuse)
plot(variogram_model, main = "Empirical Semivariogram")

# Fit a Variogram Model
vgm_model_nugget <- fit.variogram(variogram_model, 
                                  vgm(psill = 10000, model = "Nug", range = 0, add.to = 
                                      vgm(psill = 150000, model = "Sph", range = 600)))
print(vgm_model_nugget)
plot(variogram_model, vgm_model_nugget, main = "Fitted Semivariogram Model")
```

---

# **Comparing Anisotropy Across Different Angles**
```{r anisotropy-comparison, message=FALSE}
# Compute directional variograms
variogram_0 <- variogram(zinc ~ 1, meuse, alpha = 0, cutoff = 1000, width = 50)
variogram_90 <- variogram(zinc ~ 1, meuse, alpha = 90, cutoff = 1000, width = 50)
variogram_45 <- variogram(zinc ~ 1, meuse, alpha = 45, cutoff = 1000, width = 50)
variogram_135 <- variogram(zinc ~ 1, meuse, alpha = 135, cutoff = 1000, width = 50)

# Fit anisotropic models
anisotropic_model <- vgm(psill = 150000, model = "Sph", range = 800, anis = c(45, 0.5), nugget = 5000)

vgm_aniso_0 <- fit.variogram(variogram_0, model = anisotropic_model)
vgm_aniso_90 <- fit.variogram(variogram_90, model = anisotropic_model)
vgm_aniso_45 <- fit.variogram(variogram_45, model = anisotropic_model)
vgm_aniso_135 <- fit.variogram(variogram_135, model = anisotropic_model)

# Plot all directional variograms
grid.arrange(
  plot(variogram_0, vgm_aniso_0, main = "0° (East-West)"),
  plot(variogram_90, vgm_aniso_90, main = "90° (North-South)"),
  plot(variogram_45, vgm_aniso_45, main = "45° (NE-SW)"),
  plot(variogram_135, vgm_aniso_135, main = "135° (NW-SE)"),
  ncol = 2
)
```

---

# **Conclusion**
- **Semivariograms quantify spatial dependence**.
- **Key parameters**: **nugget, sill, and range**.
- **Basic statistics help interpret spatial structure**.
- **Comparing directional variograms helps detect anisotropy**.
- **Next step**: Use **kriging** for spatial interpolation.